Predicting tree failure likelihood for utility risk mitigation via artificial intelligence

### 2021-03-04 docs/manuscript/manuscript.tex

\begin{figure}[ht]
  \centering
  % \includegraphics[width=\textwidth]{raw_input_images_2}
  \begin{subfigure}[t]{.32\linewidth}\centering
    \includegraphics[width=\textwidth,angle=-90]{probable-example-1} %probable_standing dead.JPG

    \includegraphics[width=\textwidth,angle=-90]{probable-example-2} %Probable (deadwood above line hieght, vine)
    \caption{Probable}
  \end{subfigure}
    \begin{subfigure}[t]{.32\linewidth}\centering
    \includegraphics[width=\textwidth,angle=-90]{possible-example-1} % Possible (spruce, codominant and dieback)

    \includegraphics[width=\textwidth,angle=-90]{possible-example-2} % Possible (silver maple, codom weak attach, reaction wood)
    \caption{Possible}   
  \end{subfigure}
  \begin{subfigure}[t]{.32\linewidth}\centering
    \includegraphics[width=\textwidth,angle=-90]{improbable-example-1} %Improbable_3_london plane

    \includegraphics[width=\textwidth,angle=-90]{improbable-example-2}
    \caption{Improbable}
  \end{subfigure}
  \caption{Examples of training images in each of the three tree risk categories considered in this study.
  Trees in the right-hand column were categorized as improbable due to their lack of structural defects as well as good physiological health. Trees in the center column were categorized as possible due to weak branch unions and crown dieback. Trees in the left-hand column were categorized as probable because they were dead. Leaves in the bottom image in the left-hand column are from vines attached to the dead tree. Images are antialiased in this figure for greater clarity}
  \label{fig:raw_images}
\end{figure}

@@

\subsection{Classification scenarios}
In order to investigate the efficacy of an AI classifier to distinguish the failure-likelihood categories, 
we defined four classification scenarios in \autoref{tab:scenarios} for our experiments. 
Each scenario represents a unique grouping of each of the three categories, with a minimum of two derived classes in each case.
Thus, 
\begin{table}[h!]
    \centering
    \begin{tabular}{l l c}\toprule
    \bf Scenario            & \bf Description  & \bf No.\ classes\\\midrule
    \texttt{Pr\_Po\_Im}     &  \{Probable, Possible, Improbable\} & 3 \\
    \texttt{Pr\_Im}          & \{Probable, Improbable\}           & 2 \\
    \texttt{PrPo\_Im}        & \{Probable + Possible, Improbable\}  & 2 \\
    \texttt{Pr\_PoIm}        & \{Probable, Possible + Improbable\}  & 2 \\\bottomrule
    \end{tabular}
    \caption{Classification scenarios}
    \label{tab:scenarios}
\end{table}

In \autoref{tab:scenarios}, the first scenario considered each likelihood of failure category separately, as an arborist would do in practice. The second scenario considered only the highest and lowest likelihood of failure categories used in the study to clearly distinguish between categories. Since Previous research has suggested that professionals more often disagree when distinguishing between possible and probable likelihood of failure (KOESER), in the third scenario, we pooled trees in the probable and possible categories and compared them to trees in the improbable category. In the fourth scenario, we pooled trees in the possible and improbable categories and compared them to trees in the probable scenario. In practice, this scenario is less likely because arborists typically distinguish trees with an improbable likelihood of failure as those with minimal or no structural defects. It requires additional judgment to distinguish trees with probable or possible likelihood of failure because an arborist must assess the severity of structural defects, the presence of response growth, and the expected loads \cite{smiley2017best}.
\subsection{Image pre-processing and augmentation}
Data augmentation refers to the variety of methods that are employed for synthetically generating more samples in a training dataset in order to improve model performance \cite{wong2016understanding}.
Augmentation is desired, particularly in situations where the number of original observations is small, and the effectiveness of various relevant techniques in this domain has been amply demonstrated \cite{shorten2019survey}.

@@

\subsection{Hyperparameter optimization}
We optimized eight of the CNN hyperparameters using Hyperband \cite{li2018hyperband}, an efficient guided grid-search algorithm.
Twelve searches were performed for each classification scenario and image resolution combination.
Each search was conducted using 90 trials of unique hyperparameter combinations. The specified range of each parameter along with the search results are shown in \autoref{tab:opt-hyp}. For the kernel size in the first convolutional layer, we allowed for a choice between a $5\times 5$ and a $7\times 7$ kernel. The activation function in both dense layers was specified as a choice between the rectified linear unit (ReLU) function and the hyperbolic tangent ($\tanh$). The ReLU was introduced to address the so-called "vanishing gradient" problem and has been shown to improve performance in CNNs \cite{glorot2011deep}. Nevertheless, the $\tanh$ function remains a viable option, as well. The dropout rates were allowed to range from 0 to 5 in steps of 0.05, while the number of neurons or units in each dense layer varied from 32 to 512 in steps of 32. Finally, we uniformly sampled learning rates $\lambda$ for the optimizer in the $\log_{10}$ space of $[10^{-4}, 10^{-2}]$.

\begin{table}[h!]
    \centering
    \begin{tabular}{l l c}\toprule
    \bf Symbol            & \bf Definition  \\\midrule
    \texttt i     &  \ Given observation \\
    \texttt c          & \ Class\\
    \texttt s_c        & \ Class-specified score \\
    \texttt $\hat y_i$       & \ Predicted class of a given observation \\
    \texttt y_i        & \ Observed (true) class of a given observation \\
    \texttt f(s_c)        & \ Softmax activation function \\
    \texttt p_{i,c}        & \ Predicted probability that the i\textsuperscript{th} observation belongs to class c \\
    \texttt L^{CE}_i        & \ Categorical cross entropy loss function of a single observation \\
    \textttt Pr & \ Precision \\
    \textttt Re & \ Recall \\
    \textttt F_1 & \ F_1 score \\\bottomrule
    \end{tabular}
    \caption{Symbols description}
    \label{tab:symbols}
\end{table}

@@

\subsection{Model training and assessment}
The softmax activation function  $f(\cdot)$ in the output layer returns the class prediction probabilities for a given observation $i$. It is defined in terms of the class-specific score $s_c$:
\begin{equation}
    f(s_c) = \frac{e^{s_c}}{\sum\limits_{c' \in C} e^{s_{c'}}}
\end{equation}
Thus for the $i$th observation, the softmax activation returns the predicted probability $p_{i,c}$ that the $i$th observation belongs to class $c$.
The CNN is trained using a variant of the stochastic gradient algorithm, Adam \cite{kingma2017adam}.
The goal of the training procedure is to learn the optimal weights and bias terms for the CNN by minimizing a loss function. 
In this case, we use the categorical cross-entropy loss function, which for a single observation can be simply defined as:
\begin{equation}
    L_{i}^{CE} = -  \log( p_{i,c}) = - \log(f(s_c^i))
\end{equation}
Training is iteratively performed, with gradient of the loss function computed and averaged over a batch of input images.
Here, we use a batch size of 32.
The learning rate $\lambda$ of the optimization algorithm is an important hyperparameter that affects training performance.
We optimized for this in the hyperparameter search as discussed.
Furthermore, the CNN is trained over multiple passes through the entire training set.
Each such pass is referred to as an epoch.

>>>>

\begin{figure}[ht]
  \centering
  \begin{subfigure}[t]{.32\linewidth}\centering
    \includegraphics[width=\textwidth,keepaspectratio=true]{probable-example-1} %probable_standing dead.JPG

    \includegraphics[width=\textwidth,keepaspectratio=true]{probable-example-2} %Probable (deadwood above line hieght, vine)
    \caption{Probable}
  \end{subfigure}
    \begin{subfigure}[t]{.32\linewidth}\centering
    \includegraphics[width=\textwidth,keepaspectratio=true]{possible-example-1} % Possible (spruce, codominant and dieback)

    \includegraphics[width=\textwidth,keepaspectratio=true]{possible-example-2} % Possible (silver maple, codom weak attach, reaction wood)
    \caption{Possible}   
  \end{subfigure}
  \begin{subfigure}[t]{.32\linewidth}\centering
    \includegraphics[width=\textwidth,keepaspectratio=true]{improbable-example-1} %Improbable_3_london plane

    \includegraphics[width=\textwidth,keepaspectratio=true]{improbable-example-2}
    \caption{Improbable}
  \end{subfigure}
  \caption{Examples of training images in each of the three tree risk categories considered in this study.
  Trees in the right-hand column were categorized as improbable due to their lack of structural defects as well as good physiological health. Trees in the center column were categorized as possible due to weak branch unions and crown dieback. Trees in the left-hand column were categorized as probable because they were dead. Leaves in the bottom image in the left-hand column are from vines attached to the dead tree. Images are antialiased in this figure for greater clarity}
  \label{fig:raw_images}
\end{figure}

@@

\subsection{Classification scenarios}
In order to investigate the efficacy of an AI classifier to distinguish the failure-likelihood categories, 
we defined four classification scenarios in \autoref{tab:scenarios} for our experiments. 
Each scenario represents a unique grouping of each of the three categories, with a minimum of two derived classes in each case.
\begin{table}[h!]
    \centering
    \begin{tabular}{l l c}\toprule
    \bf Scenario            & \bf Description  & \bf No.\ classes\\\midrule
    \texttt{Pr\_Po\_Im}     &  \{Probable, Possible, Improbable\} & 3 \\
    \texttt{Pr\_Im}          & \{Probable, Improbable\}           & 2 \\
    \texttt{PrPo\_Im}        & \{Probable + Possible, Improbable\}  & 2 \\
    \texttt{Pr\_PoIm}        & \{Probable, Possible + Improbable\}  & 2 \\\bottomrule
    \end{tabular}
    \caption{Classification scenarios}
    \label{tab:scenarios}
\end{table}

In \autoref{tab:scenarios}, the first scenario considered each likelihood of failure category separately, as an arborist would do in practice. The second scenario considered only the highest and lowest likelihood of failure categories used in the study to clearly distinguish between categories. Since Previous research has suggested that professionals more often disagree when distinguishing between possible and probable likelihood of failure (KOESER), in the third scenario, we pooled trees in the probable and possible categories and compared them to trees in the improbable category. In the fourth scenario, we pooled trees in the possible and improbable categories and compared them to trees in the probable scenario. In practice, this scenario is less likely because arborists typically distinguish trees with an improbable likelihood of failure as those with minimal or no structural defects. It requires additional judgment to distinguish trees with probable or possible likelihood of failure because an arborist must assess the severity of structural defects, the presence of response growth, and the expected loads \cite{smiley2017best}.

\subsection{Image pre-processing and augmentation}
Data augmentation refers to the variety of methods that are employed for synthetically generating more samples in a training dataset in order to improve model performance \cite{wong2016understanding}.
Augmentation is desired, particularly in situations where the number of original observations is small, and the effectiveness of various relevant techniques in this domain has been amply demonstrated \cite{shorten2019survey}.

@@

\subsection{Hyperparameter optimization}
We optimized eight of the CNN hyperparameters using Hyperband \cite{li2018hyperband}, an efficient guided grid-search algorithm.
Twelve searches were performed for each classification scenario and image resolution combination.
Each search was conducted using 90 trials of unique hyperparameter combinations. The specified range of each parameter along with the search results are shown in \autoref{tab:opt-hyp}. For the kernel size in the first convolutional layer, we allowed for a choice between a $5\times 5$ and a $7\times 7$ kernel. The activation function in both dense layers was specified as a choice between the rectified linear unit (ReLU) function and the hyperbolic tangent ($\tanh$). The ReLU was introduced to address the so-called "vanishing gradient" problem and has been shown to improve performance in CNNs \cite{glorot2011deep}. Nevertheless, the $\tanh$ function remains a viable option, as well. The dropout rates were allowed to range from 0 to 5 in steps of 0.05, while the number of neurons or units in each dense layer varied from 32 to 512 in steps of 32. Finally, we uniformly sampled learning rates for the optimizer in the $\log_{10}$ space of $[10^{-4}, 10^{-2}]$.

\begin{table}[h!]
    \centering
    \begin{tabular}{l l c}\toprule
    \bf Symbol            & \bf Definition  \\\midrule
    $i$     &  \ Given observation \\
    $c$          & \ Class\\
    $s_c$        & \ Class-specified score \\
    $\hat y_i$       & \ Predicted class of a given observation \\
    $y_i$        & \ Observed (true) class of a given observation \\
    $f(s_c)$        & \ Softmax activation function \\
    $p_{i,c}$        & \ Predicted probability that the i\textsuperscript{th} observation belongs to class c \\
    $L^{CE}_i$        & \ Categorical cross entropy loss function of a single observation \\
    $Pr$ & \ Precision \\
    $Re$ & \ Recall \\
    $F_1$ & \ F_1 score \\\bottomrule
    \end{tabular}
    \caption{Symbols description}
    \label{tab:symbols}
\end{table}

@@

\subsection{Model training and assessment}
The softmax activation function  $f(\cdot)$ in the output layer returns the class prediction probabilities for a given observation $i$. It is defined in terms of the class-specific score $s_c$:
\begin{equation}
    f(s_c) = \frac{e^{s_c}}{\sum\limits_{c' \in C} e^{s_{c'}}}
\end{equation}
Thus for the $i$th observation, the softmax activation returns the predicted probability $p_{i,c}$ that the $i$th observation belongs to class $c$.
The CNN is trained using a variant of the stochastic gradient algorithm, Adam \cite{kingma2017adam}.
The goal of the training procedure is to learn the optimal weights and bias terms for the CNN by minimizing a loss function. 
In this case, we use the categorical cross-entropy loss function, which for a single observation can be simply defined as:
\begin{equation}
    L_{i}^{CE} = -  \log( p_{i,c}) = - \log(f(s_c^i))
\end{equation}
Training is iteratively performed, with gradient of the loss function computed and averaged over a batch of input images.
Here, we use a batch size of 32.
The learning rate  of the optimization algorithm is an important hyperparameter that affects training performance.
We optimized for this in the hyperparameter search as discussed.
Furthermore, the CNN is trained over multiple passes through the entire training set.
Each such pass is referred to as an epoch.
